Abstract

### 2022-03-30 paper/Forest_and_Water/Abstract.Rmd
update funding

---
title: Factors determining how catchments respond to forest cover change. Re-analysing global data sets.
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: DARE
    footnote: 1
  - name: Eliana Nervi
    email: eliananervi@gmail.com
    affiliation: INIA 
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: IMFIA
address:
  - code: DARE
    address: ARC Training Centre Data Analytics for Resources and Environments \& Sydney Institute of Agriculture, School of Life and Environmental Sciences.
  - code: The University of Sydney
    address: The University of Sydney, Sydney, NSW 2006, Australia
  - code: INIA
    address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria, INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: IMFIA
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la República, 11200 Montevideo, Uruguay
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover. Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data. Smaller catchment dominate the database with 42\% of the data < 1 km^2^ and 65\% of the data < 10 km^2^. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors. 
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
bibliography: forest_and_water.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

Introduction
==========================

There has been an long and on-going discussion in the hydrological literature around the impact of forests on streamflow [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017]. The historic work highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result in relation to this, is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation.

<!-- Include Filoso paper -->
For the purpose of this paper, *watershed* and *catchment* are interchangeable terms. Many of the US studies use *watershed*, while European and Australian studies use *catchment*. In particular, we retained the term "paired watershed studies" and "quasi-paired watershed studies" as this is the most common terminology, but further mostly use the term catchment.

Several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].  These studies aim to generalize the individual findings and to identify if there are global trends or relationships that can be developed. The most recent reviews [@zhang2017; @filoso2017] developed an impressive global database of catchment studies in relation to changes in streamflow due to changes in forest cover. The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets.

The conclusions of the first paper [@zhang2017] suggest that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^. While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover. Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].

The second study [@filoso2017] is a systematic review of reforestation studies (only studies in which forest cover increased). This study classified the historical research and highlighted gaps in the spatial distribution, the types of studies and the types of analysis. Their main conclusion was also that reforestation decreases streamflow, but that there were many interacting factors. For a subset of the data (37 data points) they also indicated decreasing impacts of reforestation with increasing catchment size (agreeing with @zhang2017), but they did not identify a distinct threshold and fitted a log-linear relationship. In addition, they identified that studies with shorter periods of data collection resulted in larger declines in streamflow.

A final summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017. However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model. In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model. A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017.

Encouraged by the work from @zhang2017, @filoso2017 and @zhou2015 and the large database of studies presented by these authors, we believe more can be done to add to this important discussion. In this paper, the aim is to extend the analysis of the collected data and to expand and combine the data sets.

In particular, the main method in the work by @zhang2017 is a single covariate linear regression, and in @filoso2017 the focus is mainly on classification and there is again some single covariate linear regression. As @zhang2017 points out, a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments. However, the subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum. And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact, and focussed only on forest cover increase and did not deal with forest cover removal.

As a result the objective of this paper is to 1) enhance the data set from @zhang2017 with further catchments (such as from @filoso2017) and spatial coordinates and 2) to analyse the possibility of non-linear and confounding partial effects of the different factors and variables in the data using generalised linear (GLM) and generalised additive models (GAM @wood2006).

Building on the analyses by @zhang2017 and @filoso2017, and combining their conclusions, the main hypothesis to test is that the change in streamflow is impacted by the change in forest cover. However, this change is is potentially modulated by the area under consideration (affecting the length of the flowpaths @zhou2015), the length of the study (c.f. @jackson2005; @filoso2017) and  the climate (as indicated by either E0/Pa or latitude and longitude @filoso2017; @zhou2015).

However, there could be further confounding factors, which are eluded to by @filoso2017:

* the type of analysis, i.e. paired watershed studies, modelling, time series analysis etc.

* the age of the study, assuming that historical studies might not have had the ability to measure at the accuracy that currently is available to researchers, or that more careful historical attention to detail in field studies might have been lost more recently due to reductions in research investment.

Finally, this work aims to point to further research that can expand this area of work, based on the collected data, to better understand the impact of forest cover change on streamflow.

Methods
==========

## The original data sets

The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation. 
Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.

Table 1  Summary of abbreviations of factors used in the @zhang2017 data set

|Factor | Abbreviation | Definition |
|:------------|:-------|:----------|
| forest type | CF |coniferous forest | 
|  | BF | broadleaf forest |
|  | MF |mixed forest  |
|hydrological regime|RD|rain dominated|
| |SD|snow dominated|
| climate type| EL| energy limited|
| |WL| water limited|
| |EQ| equitant|
|precipitation data type|OB|observed|
| |SG|spatial gridded|
| |MD|modelled|
|assessment technique|PWE|paired watershed experiment|
| |QPW|quasi-paired watershed experiment|
| |HM|hydrological modelling|
| |EA|elastictity analysis|
| |SH|combined use of statistical methods|
| | |and hydrographs|

While @zhang2017 use the dryness index in their analysis, and calculate the variable climate type from this index, the potential or reference evapotranspiration was not originally included as part of the published data set. In addition, dryness might mask areas where high rainfall (with potentially higher intensity rainfall) dominates the impact of high ET. In other words, high rainfall can possibly point to more infiltration excess runoff, which might be less impacted by catchment wetness condition (determined by cumulative ET).  In this paper, we do include the dryness index but did not use the climate type as a variable (as they are interchangeable).
We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.

## Additional data collection

To enhance the existing data set, this study added additional variables and cross-checked the studies with the data set from @filoso2017. In particular, we focussed on including the 37 data points related to the quantitative analysis in @filoso2017.

In addition, latitude and longitude for the center of the catchment as an approximation of its spatial location. These data were added for the different studies, mostly by using the data reported by the authors, but in some cases approximating the location of the centre of the catchment using Google Maps^TM^. In the dataset, an additional column has been added to indicate the source of the location data.

Climate more generally, and in particular the ratio of rainfall and evapotranspiration can have a significant effect on the streamflow change as represented by the dryness index, which is also highlighted by both @zhang2017 and @jackson2005. Increased evapotranspiration could lead to drier catchments, unless balanced by rainfall (such as possibly in the tropics). Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.

Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:

\begin{equation}
D = \frac{E_{0}}{Pa} (\#eq:eq1)
\end{equation}

The length of the study can be a variable influencing the change in flow [e.g. @jackson2005; @filoso2017], as for example, more mature plantations are thought to have smaller impacts on flow or regrowth might follow a "Kuczera curve" [@kuczera1987]. It is not clear if this is an effect of increased water use in growth [@vertessy2001] or due to changes in interception [@stoof2012]. Therefore, the length of the study calculate as the difference between the starting data and completion date of the different studies was extracted from the references provided by @zhang2017. The length of the study was already included in the data from @filoso2017, but these were checked against the original publications.

Several additional data points from catchment studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the accompanying data set, the main reference for the calculated change in streamflow  was generally used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011).
We also removed one data point from the analysis, which corresponds to catchment #1 (Amazon) in @zhang2017. This is because the cited reference [@roche1981] only relates to 1  and 1.5 ha paired catchment studies in French Guyana, and in which the actual change in forest cover is not recorded. Furthermore, the change in flow for catchment #76 was corrected from 600% to 157% after review of the original publication [@baker1984]. Finally, on review of all the data in @zhang2017 and @filoso2017, 29 potential duplicates were identified and flagged in the data, and not used in the analysis.

The final column in the improved data set is a "notes" column, which we added, but is not further used in the analysis. It gives context to some of the data for future research and highlights some of the discrepancies that we found between the original papers and the data in the tables from @zhang2017. This will allow future research to further sructinise our input for errors.

All the final data and analysis for this paper are located on github: [https://github.com/WillemVervoort/Forest_and_water](https://github.com/WillemVervoort/Forest_and_water/tree/publish) on the "publish" branch.

## Statistical modelling
```{r, include=FALSE}
options(tinytex.verbose = FALSE)
```
<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F, warning=FALSE}
require(tidyverse)
require(lubridate)
require(mgcv)
require(visreg)
require(patchwork)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))

Zhang_small <- read_csv("../../data/SmallCatchments_T2.csv")
Zhang_large <- read_csv("../../data/LargeCatchments_T1.csv")
new_data <- read_csv("../../data/NewCatchments_T3.csv")
```

<!--- ### change names of columns
the package`mgcv` does not seem to be happy with complex names, so probably better to make all names simple --->

```{r, echo = F}
names(Zhang_small)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_large)[3:4] <- c("Area_km2", "Pa_mm")
names(new_data)[3:4] <- c("Area_km2", "Pa_mm")
names(Zhang_small)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_large)[5:6] <- c("Forest_type", "Hydrological_regime")
names(new_data)[5:6] <- c("Forest_type", "Hydrological_regime")
names(Zhang_small)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(Zhang_large)[9:10] <- c("Precip_data_type", "Assessment_technique")
names(new_data)[9:10] <- c("Precip_data_type", "Assessment_technique")

```

<!-- combine both tables -->

```{r, echo = F, warning = F, message = F}
Zhang_small$From <- as.numeric(Zhang_small$From)
Zhang_small$To <- as.numeric(Zhang_small$To)
Zhang_all <- bind_rows(Zhang_large,Zhang_small) %>%
  mutate(dataset = "original Zhang et al data")
new_data <- new_data %>%
  mutate(dataset = "new data")
All_data <- bind_rows(Zhang_all, new_data)
```

<!--filter duplicates -->
```{r, echo = F}
All_data <- All_data %>%
  mutate(`Possible duplicate` = 
           ifelse(is.na(`Possible duplicate`)==T,0,`Possible duplicate`),
         `Possible duplicate` = as.numeric(`Possible duplicate`)) %>%
  filter(`Possible duplicate` != 1)
```

<!--now combine with E0 values -->
```{r, echo = F}
# calculate dryness index
All_data <- All_data %>%
  mutate(Dryness = E0/Pa_mm)
```

<!--remove the Amazon from the analysis -->
```{r, echo = F}
All_data <- All_data %>%
  filter(`Watershed #` != 1)
```

<!-- correct watershed #76 (Baker1984) to be 157 for DeltaQf_perc -->

```{r,echo = F}
All_data <- All_data %>%
  mutate(DeltaQf_perc = ifelse(`Watershed #` == 76,157,DeltaQf_perc))
```

To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006].

The general model tested is:

\begin{align}
\Delta Qf \% \sim &~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ 
& \sum{X_i} + \sum{s(Z_i)} + \varepsilon (\#eq:eq2)
\end{align}

Here $X_i$ are factorial variables, while $Z_i$ are continuous variables. The model intially assumes no direct interactions and all variables are additive. We will comment in this assumption in the discussion. The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover. However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. Therefore all the change in forest cover data was converted to positive values, and an additional factorial column ($sign_{forest cover}$) is included indicating whether it was a forest cover increase or decrease.

A further assumption in the model is that all continuous variables $Z_i$ (such as annual precipitation (Pa)) can have either a linear or a non-linear relationship with $\Delta Qf \%$. This means that a smooth function $s()$ can be applied to the $Z_i$ variables. For the smoothing function we applied thin plate regression splines with an additional shrinkage penalty. The result of this approach is that for high enough smoothing parameters (i.e. if the data is very "wiggly") the smooth term can be shrunk to 0 and thus will be no longer significant [@wood2006]. This is done because a highly flexible smooth term could always fit the data, but would not necessarily indicate a relevant relationship. In other words, the approach balances finding a smooth non-linear relationship for the variable against overfitting the data.

The over arching test focuses on  the change streamflow as a result of a change in forest cover being influenced by three major additional factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study. Therefore, even if these variables are insignificant in any of the applied models, we retained variables representing these three factors.

As an initial approach we  only used the data from @zhang2017 to make sure that the additional catchments added to the data set did not influence the results (this is discussed in the results). Subsequently the analysis was repeated and the additionally identified catchments were added.

Results
===========

<!-- analysis -->
## Description of the data

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The current number of catchments is the result of the removal of duplicates and our modifications and additions. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1).

```{r, echo = F, warning = F, message=F, eval =F}
require(latex2exp)
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  mutate(log10Area = log10(Area_km2)) %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, log10Area, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("log10Area" =  "log10(Area km2)", #latex2exp::TeX("$\\log_{10}({Area~km^2})$"),
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"),default = label_parsed)) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

(ref:datagraphscaption) Overview of the distribution of the data set for five of the included variables. Note that the first panel (showing the distribution of the catchment areas) indicates the distribution of the *log_10_* transformed Area (in km^2^).

```{r datagraphs, echo=F, out.width="90%", fig.cap="(ref:datagraphscaption)"}

include_graphics("./DataExploration.png")
```

```{r, echo = F}
All_data2 <- All_data %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "Decrease", "Increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))
```

```{r, eval = F, echo = F}
All_data2 %>%
  filter(is.na(Forest_Sign)==T)
```

Analysing this in more detail, the data related to forest decreases, indicate almost always a positive flow change (Figure 2). In other words, flow almost always increased. However, for increases in forest cover, this is not the case, and flow can both increase and decrease. However in both cases the variability in the reported change in flow increases with the increase in forest cover change.

```{r, echo = F, warning = F, message=F, eval =F}
# plot to show the difference
png("Increase_decrease.png", width = 2500, height = 1800, res = 300)
All_data2 %>%
  filter(is.na(Forest_Sign)==F) %>%
  ggplot(aes(DeltaF_perc_pos,DeltaQf_perc, colour = Forest_Sign)) + 
  geom_point() +  
  facet_wrap(~Forest_Sign, ncol=1, scales = "free") + 
  theme_bw() + 
    theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.text = element_text(size = rel(1.2)),
        legend.title = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5))) + 
  scale_colour_manual(name = "Forest cover direction", 
                      values = c("Decrease" = "red","Increase" = "blue")) +
  xlab("Change in forest cover %") + ylab("Change in flow (Qf) %")
dev.off()

```

```{r increasedecrease, echo =F, fig.cap="Changes in flow as a function of decreases (top) and increases (bottom) in forest cover", out.width="90%"}
include_graphics("Increase_decrease.png")
```

```{r, eval = F, echo = F}
linear_model_E0 <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign + E0, 
                           data = All_data2)
#summary(linear_model_E0)
#gam.check(linear_model_simple)

```

## The general relationship between change in forest cover and streamflow

Following @zhang2017, the first step is to investigate the percent change in flow as a linear effect of the percent change forestry and modulated by the direction of the change, either an increase in forest cover, or decrease in forest cover:

\begin{equation}
\Delta Qf \% \sim ~ \Delta \% forest~cover_{positive} + sign_{forest~cover} + \varepsilon (\#eq:eq3)
\end{equation}

```{r model1, echo = F}
linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign, 
                           data = All_data2 %>% filter(dataset == "original Zhang et al data"))
#summary(linear_model_simple)
#gam.check(linear_model_simple)

```

```{r model1b, echo = F}
linear_model_simple_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign, 
                           data = All_data2)
#summary(linear_model_simple)
r_sq <- round(summary(linear_model_simple_b)$r.sq,2)
```

```{r, echo =F}
pander(round(rbind(summary(linear_model_simple)$p.table,
                       summary(linear_model_simple_b)$p.table),2), caption="(\\#tab:tabmodel1) Summary results of the first regression model predicting change in streamflow from change in forest cover and accounting for the direction of the change. The first three rows relate to the model using the original data base from Zhang et al. (2017). The bottom three rows are the results of the model including the new data. Clearly there is no major change arising from the additional data.")
```

The overall variance explained in this model (equation \@ref(eq:eq3)) is not high with an adjusted *r^2^* of `r r_sq`, it generally supports the hypothesized relationship between the change in forest cover and the change in flow. The model suggests that for every 1\% change in forest cover, on the average, the flow changes 0.45\%. However the change in flow is different for forest cover decreases compared to forest cover increases. In fact, forest cover increases decrease flow by 29\% less than a similar decrease in forest cover causes flow to increase. So roughly speaking, a 1\% forest cover increase on the average decreases flow by $(1 - 0.29)*0.45\%$, while a the percentage forest cover decrease will increase flow by 0.45\%.

```{r, eval = F, echo = F}
png("residual_plot_model1.png", width = 2500, height = 1800, res = 300)
par(mfrow=c(2,2))
gam.check(linear_model_simple)
par(mfrow=c(1,1))
dev.off()
```

```{r gamcheck, echo =F, fig.cap="Residual plots for the first simple regression model indicating a slightly fat-tailed residual distribution", out.width="90%"}
include_graphics("residual_plot_model1.png")
```

Of importance here is to highlight the residuals of this regression (equation \@ref(eq:eq3) and Figure \@ref(fig:gamcheck)). These are approximately normal, although there is still significant skew on the upper and lower parts of the distribution (Figure \@ref(fig:gamcheck)). In other words, the distribution of the residuals is somewhat fat-tailed. We will discuss this later.

Including the data from some of the newly identified studies indicates that this mainly strengthens the difference between the forest cover increases and decreases (Table \@ref(tab:tabmodel1)), and the result indicate a reduction in the mean decrease in flow as a result of forest cover change if the new data is included. Adding the new data does not change the outcome much (apart from the magnitudes of the coefficients), which is expected as the number of added catchments is small relative to the total @zhang2017 data set. But this also means that our re-analysis of the data can be directly compared to the original study.

However, it is clear from the lack of explaining power for the model, that there could be confounding factors, as alluded to in the methods. The obvious ones being catchment dryness and area (following @zhang2017), which we will analyse later.

## The effect of location on the globe
<!-- include Lat and Long (later E0) -->

```{r globalmap, fig.cap="Distribution of included catchments across the globe based on reported or estimated latitude and longitude", out.width="90%", echo=F}

include_graphics("../../data/FAOET0data_final_2022.png")
```

Latitude and longitude might reveal strong spatial clustering of the studies, or latitude and longitude might indicate strong climate gradients. As the global map (Figure \@ref(fig:globalmap)) shows, the distribution of case study catchments covers multiple continents and shows some distinct clustering in parts of the world. Of interest is whether the spatial clustering also indicates a difference in response to forest cover change:

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ &  Latitude + Longitude + \varepsilon (\#eq:eq4)
\end{align}

```{r model3, echo = F}
model3_includeLatLong <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Latitude + Longitude, data = All_data2)
#summary(model3_includeLatLong)
#gam.check(model3_includeLatLong)
```

```{r, echo =F}
pander(round(summary(model3_includeLatLong)$p.table,2), caption= "(\\#tab:out-model3) Results of the model based on the complete dataset and including Latitude and Longitude")
```

There appears to be no significant gradient in either latitude or longitude (Table \@ref(tab:out-model3)), suggesting that the distribution of the catchments across the globe has little influence. The total explaining power of the model is still low with an adjusted *r^2^* of `r round(summary(model3_includeLatLong)$r.sq,2)` suggesting further factors influencing the change in streamflow that are currently not included in the model.

## Impact of climate

While latitude and longitude might hint at climatic gradients (for example a change in response related to tropical or sub tropical belts), annual rainfall and potential evapotranspiration might give a better indication.
Potential evapotranspiration (*E0*) by itself was  not significant in the. Initially, we also tested models using only the annual average precipitation (*Pa (mm)*), but  interactions between precipitation and evapotranspiraton might be captured by the dryness index. Both dryness index  and *Pa (mm)* were initially analysed as a key variables, but these indicated that these two variables were essentially interchangeable. As a result only the dryness index was retained as a climate indicator to align with the earlier work by @zhang2017. Given that Latitude and Longitude were not significant, we dropped these from the model.

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & Dryness  + \varepsilon (\#eq:eq5)
\end{align}

<!--\newpage-->

<!-- include Dryness index -->

```{r model4, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model4_dryness <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness, data = All_data2)
#summary(model4_dryness)
#gam.check(model4_dryness)
```

```{r, echo =F}
pander(round(summary(model4_dryness)$p.table,2), caption = "(\\#tab:out-model4) Results of the model including the dryness index")
```

<!-- running PCA, this is used later -->

```{r testing_PCA, eval = F, echo = F, warning = F}
require(FactoMineR)
require(factoextra)
 res.pca <- PCA(All_data2 %>% mutate(length = To - From) %>% select(`Watershed #`, `Watershed name`, Area_km2, DeltaF_perc, DeltaQf_perc,  Dryness,  From, length, Latitude, Longitude), scale.unit = TRUE, ncp = 4, quali.sup = 1:2, graph=FALSE)
 # scree.plot <- fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 41))
 #  scree.plot
 #  
 #   var.plot <- fviz_pca_var(res.pca, axes = c(1, 2), col.var = "black")
 #  var.plot
 png("PCA_biplot_Alldata.png", width = 800, height = 800)
  biplot_test <- fviz_pca_biplot(res.pca, axes = c(1, 2), repel = TRUE,
                                                      col.var = "darkblue", # Variables color
                            label="var", #"var" 
                            col.ind = "red", # Individuals color
                            title = "",labelsize = 8, pointsize = 4, arrowsize = 1) +
        theme(text = element_text(size = 14, face = "bold"),
              axis.title = element_text(size = rel(2), face = "bold"),
        axis.text = element_text(size = rel(2), face = "bold"))
 print(biplot_test)
 dev.off()

```

Similar to *E0* or *Pa_mm*, the results from this model (equation \@ref(eq:eq5) and Table \@ref(tab:out-model4)) interestingly indicate no impact of dryness on the change in streamflow as a function of the change in forest cover change. This might seem suprising in light of earlier reported results [@zhang2017; @filoso2017]. In this case, the evidence is highly doubtful (p = `r round(summary(model4_dryness)$p.table[4,4],2)`). However, it is very well possible that there is a further interaction in the data with other variables or unknown variables that this simpler version of the model cannot identify. This is partly evidenced by the fact that the overall variance explained is still low, with an adjusted *r^2^* of `r round(summary(model4_dryness)$r.sq,2)`. As indicated in the methods, we retain Dryness in further models as an indicator of climate for the catchments.

```{r, echo = F}
pander(All_data2 %>% filter(Dryness > 4) %>% select(`Watershed #`,Latitude, Longitude,`Watershed name`) %>%
         rename("Number" = `Watershed #` , `Catchment name` = `Watershed name`), caption = "(\\#tab:drytable) catchments for which the dryness index > 4")
```

There are also possible issues with the data, as a few of the catchments have Dryness values that are very large (> 4) and these values have high leverage in the data, affecting the residual distribution. These catchments are listed in Table \@ref(tab:drytable).

## Is there a distinct effect of area?

The second major variable is the effect of area on the change in flow, following the analysis by @zhang2017 and @filoso2017. Given the highly skewed distribution of the catchment areas (Figure \@ref(fig:datagraphs)), a log base 10 transformation was applied to the variable *Area (km^2^)*.

\begin{align}
\Delta Qf \% \sim ~ & \Delta \% forest~cover_{positive} + sign_{forest~cover} + \notag \\ & log10(Area ~(km^2)) + Dryness + \varepsilon (\#eq:eq6)
\end{align}

```{r modelArea, echo = F}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)
model_Area <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness + log10(Area_km2), data = All_data2)
#summary(model_Area)
#gam.check(model_Area)
```

```{r, echo =F}
pander(round(summary(model_Area)$p.table,2), caption = "(\\#tab:out-modelArea) Results of the model including Area and the dryness index")
```

```{r modelArea_b, echo = F, eval =T}
#Zhang_all2 <- Zhang_all2 %>%
#  mutate(length = To - From)

model_Area_b <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign +  
                        Dryness*log10(Area_km2), data = All_data2)
#summary(model_Area_b)
#gam.check(model_Area)
```

The results of this model (Equation \@ref(eq:eq6)) indicate there is at least some evidence (p = `r round(summary(model_Area)$p.table[5,4],2)`) that there is a reduction in the effect of forest cover change on streamflow related to log10 (Area (km^2^)) (Table \@ref(tab:out-modelArea)). In fact, the results suggests that for every additional 10 km^2^ in catchment size the mean change in flow reduces by `r abs(round(summary(model_Area)$p.table[5,1],1))`\%. Another interesting fact to note is that with the inclusion of Area (km^2^) as a variable in the model, the effect of Dryness becomes slightly more important, possibly suggesting an interaction between Dryness and Area. Including the interaction Dryness*log10(Area (km^2^)) in the model (Table \@ref(tab:out-modelArea-int)) results in the increased evidence (p = `r round(summary(model_Area_b)$p.table[3,4],2)`) that Dryness affects the change in flow caused by changes in forest cover and that the effect of Area is only important (p = `r round(summary(model_Area_b)$p.table[5,4],2)`)) as an interaction with Dryness.

```{r, echo =F}
pander(round(summary(model_Area_b)$p.table,2), caption = "(\\#tab:out-modelArea-int) Results of the model including an interaction between Area and the dryness index")
```

```{r, eval = F, echo = F}
All_data2 %>%
  ggplot(aes(log10(Area_km2),Dryness, size = DeltaF_perc_pos)) + geom_point(colour = "blue", alpha =0.5)
summary(lm(Dryness~log10(Area_km2),data = All_data2))
```

## Are some of the variables possibly non-linear?

The work by @filoso2017 and earlier by @jackson2005 has indicated that the length of the study might influence the response. This links to the idea from @kuczera1987 that the effect of logging or deforestation or reforestation reduces with the length of time post intervention (see also @jackson2005). In addition to adding *length* (being the difference between the reported start date and end date of data collection in the specific study) as a variable, two other continuous variables (*Dryness* and *Area*) were considered non-linear. As a result a shrinkage smoothing spline [@wood2006] was applied to these variables.

>>>>

---
title: "Abstract"
output: pdf_document
---

  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation/de-forestation and streamflow. However, the past approaches in the literature are variable and can be substantially improved in statistical rigour, and indicate different confounding factors on the impact of forestation. The data for these three papers were reviewed, combined and re-analysed to answer the following new and older questions: 1) How is streamflow impacted by the change in forest cover as a function of catchment area; 2) how is this relationship conditioned by the length of the study, and climate; and 3) are there other possible variables that impact the observed change in streamflow? Generalised additive models were used to run flexible regressions including multiple variables. 
  Changes in forest cover cause changes in streamflow, however this change is different between deforestation and reforestation, and strongly affected by climate, with drier climates indicating larger changes in streamflow. Removal of forest cover causes a 32\% greater change in flow relative to increasing forest cover. Area of the catchment only affects the change in streamflow after log transformation, due to high skew in the data. Smaller catchment dominate the database with 42\% of the data < 1 km^2^ and 65\% of the data < 10 km^2^. Length of the study and initial year of the study did not affect the change in flow, in contrast to other reported studies. Despite these findings, overall explained variance (38\%) of the regression model is low due the quality of the inputs and additional unknown confounding factors.
